Web Processing - Standardised GIS Analyses for Cable Route Planning

### 2023-03-08 paper/discussion.tex
grammar + change to en_GB

% !TeX spellcheck = en_US
As shown, the need for computation time increases with the power of two with the resolution. 
Similarly, the usage of main memory increases. 
Which in tern limits the processable data points and resolution.
On the error only scales linear with the resolution. 
Hence, there is a diminishing return of smaller errors, compared to compute time and resources used.

@@

As the original least cost path QGIS plug-in the used algorithm stops the least aggregation of the costs after the final end point has been found.
This early stopping might result in suboptimal paths around the end points for some edge cases, where the connection via another neighbor might be more optimal.

@@

In this work, the single cost raster layers aggregated with the maximum function. 
Another possible aggregation function is the sum.
Each aggregation function can be justified, by a different interpretation of the costs and its scale.
The reasoning of selection the maximum function is, limiting the maximum costs.
When the \textit{prohibited} level is used as the highest level.
Then higher levels are \textit{prohibited} within the concept.
This an area even more \textit{prohibited}, when it is included in two different rules of the level?
On the other side, This aggregation is unable to differentiate between these nuances, as  the sum function is.
This way, one can differential between, different sublevels.
One consequence is, that a small mediation between high and low single cost raster, in the final cost raster is possible.

The fact, that all touched False raster lead to more similar results compared to high resolution raster, is due to the fact, that the default level is relatively low.
In cause the default level would rise, the effect probably would flip for low resolution raster.
For high resolution raster, this effect would still hold true, because the fact, that the raster pixel center is used for sampling, reflects the original geometry better.

>>>>

% !TeX spellcheck = en_GB
As shown, the need for computation time increases with the power of two with the resolution. 
Similarly, the usage of main memory increases. 
Which in tern limits the processable data points and resolution.
On the error only scales linear with the resolution. 
Hence, there is a diminishing return of smaller errors, compared to compute time and resources used.

@@

As the original least cost path QGIS plug-in the used algorithm stops the least aggregation of the costs after the final end point has been found.
This early stopping might result in suboptimal paths around the end points for some edge cases, where the connection via another neighbour might be more optimal.

@@

In this work, the single cost raster layers aggregated with the maximum function. 
Another possible aggregation function is the sum.
Each aggregation function can be justified, by a different interpretation of the costs and its scale.
The reasoning of selection the maximum function is, limiting the maximum costs.
When the \textit{prohibited} level is used as the highest level.
Then higher levels are \textit{prohibited} within the concept.
This an area even more \textit{prohibited}, when it is included in two different rules of the level?
On the other side, This aggregation is unable to differentiate between these nuances, as  the sum function is.
This way, one can differential between, different sub-levels.
One consequence is, that a small mediation between high and low single cost raster, in the final cost raster is possible.

The fact, that all touched False raster lead to more similar results compared to high resolution raster, is due to the fact, that the default level is relatively low.
In cause the default level would rise, the effect probably would flip for low resolution raster.
For high resolution raster, this effect would still hold true, because the fact, that the raster pixel centre is used for sampling, reflects the original geometry better.

### 2023-03-08 paper/methods.tex
grammar + change to en_GB

% !TeX spellcheck = en_US

We retrieve a set of different spacial data-sets from  public sources as a basis to create the cost raster.
Field of study are the counties of Cuxhaven and Osterholz in the state of Lower Saxony, Germany.
Areas protected by different European and National conservation laws are provided by the German Environment Agency
as \acrfull{wfs}~\cite{noauthor_schutzgebiete_2015}.

The nationwide land coverage (ATKIS) with a scale of 1:250000 are provided by the Federal Agency for Cartography
and Geodesy~\cite{noauthor_digitales_2021}.
The nationwide power grid (tags: 'power': line) has been retrieved via OpenStreetMap~\cite{boeing_osmnx_2017}.
Local data as houses at Level of Detail 1 are offered by the State Office for Geoinformation and Land Surveying of
Lower Saxony~\cite{noauthor_opengeodatani_2022}.
In addition, local planning geodata for the land usage are taken
from 'Metropolplaner' (Planing data Lower Saxony \& Bremen)~\cite{noauthor_metropolplaner_2022}.

PyWPS~\cite{noauthor_welcome_2016} is used to provide the least cost path algorithm as a \acrfull{wps} in combination with flask~\cite{noauthor_flask_nodate}.
As client Birdy~\cite{noauthor_birdy_nodate} connects to the \acrshort{wps}, sending the cost raster and receiving the resulting least cost path.
The initial implementation of the least cost path algorithm is based on the implementation for the QGIS-Plugin
'Least Cost Path'~\cite{noauthor_leastcostpathdijkstra_algorithmpy_2022} in version 1.0.

The different layers from the different entities are optionally filtered, buffered and then rasterized.
Filtering the layers of the files for special attributes enables to further differentiate.
For examples makes it possible to differentiate between heath and uncultivated land in the land coverage.
Adding a buffer can be used either used to convert a line objects as power grids and streets into a polygon with the
correct physical width, or to add minimum distance from an existing of planed area to the new power grid.
Each of theses rasterized layers are given a weight, or cost that expresses the cost of using land covered by this layer.
The costs of all layers is aggregated with the maximum function.
Thus, an area covered by multiple layers is uniformly used with the highest costs.
Any place in the study area, that is not covered by any layer and thus does not have a weight yet, is given the default cost.

The costs have been grouped into different levels~(see table~\ref{tab:1}) starting from \textit{preferential} areas with
very low costs, via \textit{no restriction}s, which is the default, used when no other layers are covering the local area,
to \textit{restricted}, \textit{strongly restricted} and \textit{prohibited} areas with high costs.
These higher costs resemble the degree how much a place of this cost should be avoided, while routing the path.
The ratio of the higher costs to the lower costs directly translates into the additional diversion in pixels
the algorithm is willing to go, for avoiding an area of this high costs.
Thus, as \textit{prohibited} areas describe a legal obligation, not to use these areas or only to the utmost minimum,
the weight that resembles the costs for these types of areas, has to be especially high.\\

All these layers are provided as vectors.
While the least cost algorithm use raster data.
The rasterization transforms a vector into a raster.
The rasterization can be executed in two different ways.
In both ways, the rasterization can be imaged, as the old vector is layered over the new raster grid with the new
given resolution and the new affine transformation and the coordinate reference system of the vector.
Both rasterization techniques differ in the selection of the pixel, that describe the original polygon.
The pixels can be selected , if either the center of that pixel is overlaid
by the geo-object, or any part of the pixel is overlaid.
The version with any part overlaid is called all touched True.
The version where a overlaid pixel center is require is called all touched False.
All touched False is considered the default (see figure~\ref{fig:alltouched}).
\begin{figure}[!ht]
	\centering

@@

The completed list of layers and their processing, can be found in \textbf{Supplement S1}.

All three steps of the generation of the least cost path: generation of the cost raster, aggregation and back tracking is shown with an example for a starting cost raster of 50~m resolution and a rasterization with all touched False (see \ref{fig:costs2path}).

The chosen implementation applies early stopping.
Therefore, the costs for points that are not needed to try to connect to the end point are not aggregated.
After an aggregated cost for every end point has been found, the aggregation stops and the back tracking starts.
Because the end point ends at a power transformer, which is a building type, the paths ends at in a \textit{Prohibited} Area.
Therefore, areas even farther away from the starting point have been explored first.

@@

%% The aggregation is implemented in a priority queue. 
%% Hence a point with a lower cost is evaluated earlier as a starting position to evaluate the cost of its neighbors, than a point with higher costs.

>>>>

% !TeX spellcheck = en_GB

We retrieve a set of different spacial data-sets from  public sources as a basis for creating the cost raster.
The study area are the counties of Cuxhaven and Osterholz in the state of Lower Saxony, Germany.
Areas protected by different European and National conservation laws are provided by the German Environment Agency
as \acrfull{wfs}~\cite{noauthor_schutzgebiete_2015}.

The nationwide land coverage (ATKIS) with a scale of 1:250000 are provided by the Federal Agency for Cartography
and Geodesy~\cite{noauthor_digitales_2021}.
The national power grid (tags: 'power': line) has been retrieved via OpenStreetMap~\cite{boeing_osmnx_2017}.
Local data as houses at Level of Detail 1 are provided by the State Office for Geoinformation and Land Surveying of
Lower Saxony~\cite{noauthor_opengeodatani_2022}.
In addition, local planning geodata for the land use are taken
from 'Metropolplaner' (Planning data Lower Saxony \& Bremen)~\cite{noauthor_metropolplaner_2022}.

PyWPS~\cite{noauthor_welcome_2016} is used to provide the least cost path algorithm as a \acrfull{wps} in combination with flask~\cite{noauthor_flask_nodate}.
As client Birdy~\cite{noauthor_birdy_nodate} connects to the \acrshort{wps}, sends the cost raster and receives the resulting least cost path.
The initial implementation of the Least Cost Path algorithm is based on the implementation for the QGIS-Plugin
'Least Cost Path'~\cite{noauthor_leastcostpathdijkstra_algorithmpy_2022} in version 1.0.

The different layers of the different entities are optionally filtered, buffered and then rasterised.
Filtering the layers of the files for special attributes enables further differentiation.
For examples, it possible to differentiate between heath and uncultivated land in the land cover.
Adding a buffer can be used either used to convert a line objects such as power grids and streets into a polygon with the
correct physical width, or to add minimum distance from an existing of planed area to the new power grid.
Each of theses rasterised layers are given a weight, or cost which expresses the cost of using land covered by this layer.
The costs of all layers is aggregated with the maximum function.
Thus, an area covered by several layers is uniformly used with the highest cost.
Any place in the study area, that is not covered by any layer and thus does not yet have a weight, is given the default cost.

The costs have been grouped into different levels~(see table~\ref{tab:1}), starting from \textit{preferential} areas with
very low costs, via \textit{no restriction}s, which is the default, used when no other layers are covering the local area,
to \textit{restricted}, \textit{strongly restricted} and \textit{prohibited} areas with high costs.
These higher costs resemble the degree to which a place with this cost should be avoided, while routing the path.
The ratio of the higher costs to the lower costs directly translates into the additional detour in pixels
the algorithm is willing to go.
Thus, as \textit{prohibited} areas describe a legal obligation, not to use these areas or only to the utmost minimum,
the weight that resembles the costs for these types of areas, has to be especially high.\\

All these layers are provided as vectors.
The Least Cost Path algorithm uses raster data.
Rasterisation transforms a vector into a raster.
Rasterisation can be done in two different ways.
In both ways, the rasterisation can be imaged, as the old vector is superimposed on the new raster grid with the new
given resolution and the new affine transformation and the coordinate reference system of the vector.
Both rasterisation techniques differ in the selection of the pixel, that describe the original polygon.
The pixels can be selected , if either the centre of that pixel is overlapped
by the geo-object, or any part of the pixel is overlaid.
The version with any part overlapped is called all touched True.
The version where a overlapped pixel centre is require is called all touched False.
All touched False is considered the default (see figure~\ref{fig:alltouched}).
\begin{figure}[!ht]
	\centering

@@

The complete list of layers and their processing, can be found in \textbf{Supplement S1}.

All three steps of the generation of the least cost path: generation of the cost raster, aggregation and backtracking is shown with an example for a starting cost raster of 50~m resolution and a rasterisation with all touched False (see \ref{fig:costs2path}).

The chosen implementation applies early stopping.
Therefore, the costs for points that are not needed to try to connect to the end point are not aggregated.
After finding an aggregated cost for every end point, the aggregation stops and the backtracking starts.
Because the path ends at a power transformer, which is a building type, the paths ends at in a \textit{Prohibited} area.
Therefore, areas even further away from the starting point have been explored first.

@@

%% Aggregation is implemented in a priority queue. 
%% Hence a point with a lower cost is evaluated earlier as a starting position to evaluate the cost of its neighbours, than a point with higher cost.

### 2023-03-08 paper/paper-CableRoute.tex
grammar + change to en_GB

% !TeX spellcheck = en_US

@@

	\begin{abstract}
		The least cost paths for a power line route from a power station is computed to a power station further inland.
		Down sampling and clipping as studied as techniques to speed-up the computation.
		The path from clipped cost raster often matches the original path.
		The speed-up of the clipping can not be known before hand.
	\end{abstract}

@@

	The German power system adds a lot of renewable energy sources in the north, because wind energy reaches highest efficiency with high wind speeds.
	At the same time old power plants e.g.\ nuclear, hard coal and lignite are going out off service~\cite{eser-frey_analyzing_2012}.
	These older power plants were mainly situated in southern and central Germany. 
	The power sink, which is industrial and private demand does not shift northwards.
	Therefore, the renewable energies have to be transported from north to south which increases the congestion in the power grid. 
	The amount of offshore wind power, that the German energy system can use, can be greatly increased by adding new power lines~\cite{leuthold_nodal_2005}.

	But for building power lines, finding the shortest path is not sufficient.
	Additional parameters also have to be taken into consideration.
	As the steepness of a road or the soil, play an important role for the building cost of a road or pipeline~\cite{suleiman_optimal_2015}.
	For planing the additional routes for a power grid, further aspects as legal regulations and acceptance
	by the local population have to be taken in consideration.
	Also technical aspects, as the effects on the grid stability are further points take into consideration~\cite{schafer_understanding_2022}.

	Due to the increasing need for renewable energy wind offshore wind turbines supply 5.5~\% annual percentage in 2020 of the German energy mix~\cite{noauthor_nettostromerzeugung_2021}.

	Other methods as the \acrfull{MCDA}  consider further aspects as stakeholder, experts, public and decision maker~\cite{bertsch_participatory_2016} (see section~\ref{sec:related-works}).

	For planning a possible route for a power line the least cost path algorithm is applied.
	The least cost path algorithm is a Dijkstra algorithm~\cite{dijkstra_note_1959} applied on a raster map.
	The vertices of the graph are the pixel centers, that are connected to the eight neighboring pixel centers via the edges.
	This enables to find routes on graphs that are not predefined like road notworks.
	The weights of the edges are the local costs to travel from one pixel center to the neighboring pixel center.
	The costs might be physical costs, as the local slope, but also can be composed on other factors as the acceptance rate for the land usage of a given class.
	The least cost path algorithm consists of at least two sequential steps.
	1) The first step is to aggregate the costs of traveling from the starting point to a given set of end points.
	This step generates the aggregated cost raster of traveling from the starting point to any point of the cost raster.
	2) In the next step the back tracking, the route of the actual least cost path is calculated.
	For each ending point the path via the lowest cost neighbor is taken until arriving at the starting point.

	Some implementations switch the role of start and ending points, so that either many starting points and one end point, or one end point and many endpoints can be used.
	In some implementation there is a extra step between 1) and 2) that calculates cost-weight direction raster, that encodes the direction of the shortest
	path to the starting point as integer values.

@@

	Older models are e.g.\ are optimizing the well fare, economic modeling is not sufficient for planning modern routes, as the aspect of environmental sustainability, security of supply and the public acceptance more and more play an important role~\cite{hauff_gesellschaftliche_2011} in modern planning.

@@

	The different models of the power system, that are used in the expert system span from not modeling the grid at all to models that includes Kirchhoffs laws~\cite{bertsch_participatory_2016}.
	The calculation of a power system is complex, as changing one edge will change the flow in all parts of grid and the actual grid data are confidential~\cite{bertsch_participatory_2016}.
	Some models include not only the grid of a nation alone, but consider neighboring states or the overall European power system~\cite{dietrich_will_2010}.

>>>>

% !TeX spellcheck = en_GB

@@

	\begin{abstract}
		The least cost paths for a power line from a power station at the cost to a power station further inland are calculated.
		Down-sampling and clipping are investigated as techniques to speed up the calculation.
		The path from the clipped cost raster is often the same as the original path.
		The speed-up of the clipping can not be known in advance.
	\end{abstract}

@@

	The German power system is adding a lot of renewable energy sources in the north, because wind energy reaches its highest efficiency with high wind speeds.
	At the same time old power plants e.g.\ nuclear, hard coal and lignite are being phased out~\cite{eser-frey_analyzing_2012}.
	These older power plants were mainly located in southern and central Germany. 
	The energy sink,  industrial and private demand, is not shifting north.
	Therefore, the renewable energy has to be transported from north to south which increases the congestion in the power grid. 
	The amount of offshore wind power, that the German energy system can use, can be greatly increased by adding new power lines~\cite{leuthold_nodal_2005}.

	However, it is not enough to find the shortest route when building new power lines.
	Other parameters have to be taken into consideration.
	As the steepness of a road or the soil, play an important role for the building cost of a road or pipeline~\cite{suleiman_optimal_2015}.
	When planing the additional routes for a power grid, further aspects such as legal regulations and acceptance
	by the local population have to be taken in consideration.
	Also technical aspects, as the effects on the grid stability are further points take into consideration~\cite{schafer_understanding_2022}.

	Due to the increasing demand for renewable energy wind offshore wind turbines supply 5.5~\% annual percentage in 2020 of the German energy mix~\cite{noauthor_nettostromerzeugung_2021}.

	Other methods such as the \acrfull{MCDA}  consider further aspects as stakeholder, experts, public and decision maker~\cite{bertsch_participatory_2016} (see section~\ref{sec:related-works}).

	The least cost path algorithm is used to plan a possible route for a power line.
	The least cost path algorithm is a Dijkstra algorithm~\cite{dijkstra_note_1959} applied on a raster map.
	The vertices of the graph are the pixel centres, that are connected to the eight neighbouring pixel centres via the edges.
	This makes it possible to find routes on graphs that are not predefined, such as road networks.
	The weights of the edges are the local costs of travel from one pixel center to the neighbouring pixel center.
	The costs can be physical costs, such as the local slope, but also can be composed of other factors as the acceptance rate for the land usage of a given class.
	The Least Cost Path algorithm consists of at least two successive steps.
	1) The first step is to aggregate the costs of travelling from the starting point to a given set of end points.
	This step generates the aggregated cost raster of travelling from the starting point to any point of the cost raster.
	2) In the next step the backtracking, the route of the actual Least Cost Path is calculated.
	For each end point the path via the lowest cost neighbour is taken until the starting point is reached.

	Some implementations switch the role of start and ending points, so that either many start points and one single end point, or one single end point and many end points can be used.
	In some implementation there is an extra step between 1) and 2) that calculates cost-weight direction raster, that encodes the direction of the shortest
	path to the starting point as integer values.

@@

	Older models are e.g.\ are optimizing the well fare, economic modelling is not sufficient for planning modern routes, as the aspect of environmental sustainability, security of supply and the public acceptance more and more play an important role~\cite{hauff_gesellschaftliche_2011} in modern planning.

@@

	The different models of the power system, that are used in the expert system span from not modelling the grid at all to models that includes Kirchhoffs laws~\cite{bertsch_participatory_2016}.
	The calculation of a power system is complex, as changing one edge will change the flow in all parts of grid and the actual grid data are confidential~\cite{bertsch_participatory_2016}.
	Some models include not only the grid of a nation alone, but consider neighbouring states or the overall European power system~\cite{dietrich_will_2010}.

### 2023-03-08 paper/results.tex
grammar + change to en_GB

In contrast when the resolution the is much less, than the size of the object the described behavior changes.
On one hand, while the area, of the pixel is increased for all touched True, also the area for which the cost is overestimated increases.
On the other hand while the pixel size for all touched  False increases, not only the over estimated area increases for that pixel as describes for all touched true, but in addition the object is less probable situated in the center of the pixel.
The consequence of the pixel is, that will decreasing resolution that object is not included in the rasterziation.
Hence, hence for all touched false lower resolution  leads ot a loss of information.
Because the default is relative small compared the over effect is a underestimation of the costs.
The figure~\ref{fig:costs_100m} shows for the resolution of 100 m, that while larger are still included in the map they appear to be much larger.
On the contrary smaller objects might not be included.
Objects that are small on in one dimension as streets will be included in a stochastic manner.
Described by the likely of that a object of that sized overlaps with the pixel center.
\begin{figure}
	\centering

@@

The final least costs paths should between the least cost paths of the lower resolution, with a tendency to be nearer to the paths resulted by the all touched false rasterization.
The first step is to optimize the calculation speed is, only to calculated the least cost paths for this smaller area.
Another method, is to improve the prediction of the medium resolution itself and thus reduce the need for a computation in higher resolution.
\subsubsection{Compare least cost paths, for overlay of both rasterizations}

@@

The optimal ratio of overlaying all touched false and all touched true cost raster for 10~m resolution is estimated via the resulting least cost path. 
The mean distance of least cost paths for different ratio has been estimated to the path from the all touched false raster of the next higher (5~m) resolution. 
Table~\ref{tab:4} shows, that distance decreases with the higher ratio from 1:1, 2:1, 4:1 and than increases with higher ratio 8:1, 16:1 and so on.
In addition, the distances to the original paths from the all touched false and all touched true raster change with the applied ratio of the overlayed raster.
So that, a lower ratio increases the similarity to the path of the all touched true raster, the similarity to the paths from the all touched false raster increases with the ratio.

@@

This enabled the possibility  to run a 2.5~m resolution cost raster and clip it to the extent of the polygon. 
This clipped 2.5~m raster for all touched true did only slightly change the path. 
While the all touched False raster, leads to a totally new segment at the end of the path. 
Due to the small resolution, that a small path with the extent of about 2.5 m width became passable.
This small path is a power line next to road between which build a cone in a protected landscape area.
The street and the protected landscape area are both restricted areas, while the power line is preferential.
This showing, that the way the cost raster is created in the first place can play a vital role, in the end result.
So that a nuance, can cause a detour.
When this behavior occurs, the polygon can not grantee to include the least cost path.
Therefore this polygon should be overlayed with a polygon around the shortest path.

>>>>

In contrast when the resolution the is much less, than the size of the object the described behavior changes.
On one hand, while the area, of the pixel is increased for all touched True, also the area for which the cost is overestimated increases.
On the other hand while the pixel size for all touched  False increases, not only the over estimated area increases for that pixel as describes for all touched true, but in addition the object is less probable situated in the centre of the pixel.
The consequence of the pixel is, that will decreasing resolution that object is not included in the rasterisation.
Hence, hence for all touched false lower resolution  leads ot a loss of information.
Because the default is relative small compared the over effect is a underestimation of the costs.
The figure~\ref{fig:costs_100m} shows for the resolution of 100 m, that while larger are still included in the map they appear to be much larger.
On the contrary smaller objects might not be included.
Objects that are small on in one dimension as streets will be included in a stochastic manner.
Described by the likely of that a object of that sized overlaps with the pixel centre.
\begin{figure}
	\centering

@@

The final least costs paths should between the least cost paths of the lower resolution, with a tendency to be nearer to the paths resulted by the all touched false rasterization.
The first step is to optimize the calculation speed is, only to calculated the least cost paths for this smaller area.
Another method, is to improve the prediction of the medium resolution itself and thus reduce the need for a computation in higher resolution.
\subsubsection{Compare least cost paths, for overlay of both rasterisations}

@@

The optimal ratio of overlaying all touched false and all touched true cost raster for 10~m resolution is estimated via the resulting least cost path. 
The mean distance of least cost paths for different ratio has been estimated to the path from the all touched false raster of the next higher (5~m) resolution. 
Table~\ref{tab:4} shows, that distance decreases with the higher ratio from 1:1, 2:1, 4:1 and than increases with higher ratio 8:1, 16:1 and so on.
In addition, the distances to the original paths from the all touched false and all touched true raster change with the applied ratio of the superimposed raster.
So that, a lower ratio increases the similarity to the path of the all touched true raster, the similarity to the paths from the all touched false raster increases with the ratio.

@@

This enabled the possibility  to run a 2.5~m resolution cost raster and clip it to the extent of the polygon. 
This clipped 2.5~m raster for all touched true did only slightly change the path. 
While the all touched False raster, leads to a totally new segment at the end of the path. 
Due to the small resolution, that a small path with the extent of about 2.5 m width became passable.
This small path is a power line next to road between which build a cone in a protected landscape area.
The street and the protected landscape area are both restricted areas, while the power line is preferential.
This showing, that the way the cost raster is created in the first place can play a vital role, in the end result.
So that a nuance, can cause a detour.
When this behaviour occurs, the polygon can not grantee to include the least cost path.
Therefore this polygon should be overlapped with a polygon around the shortest path.
